A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-10 podc_paper/algorithm.tex
minor edits

As mentioned above, we are free to order operations within a block however we like.
We order the enqueues and dequeues separately, and put the 
operations propagated from the left child before the operations from the right child.
More formally, we inductively define sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
represented by a block $B$.
If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is that operation
and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from 
$v$'s left child
and $B^R_1,\ldots,B^R_r$ from $v$'s right child, then $B$'s operation sequences are defined by the concatenations 
\begin{eqnarray}
E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)\nonumber\\
D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
\end{eqnarray}
We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.

@@

The information stored in a \block\ depends on whether it is in an internal node or a leaf.
A block in a leaf represents a single enqueue or dequeue.  The \fld{element} field stores the value
enqueued for an enqueue operation, or \nl\ if the operation is a dequeue.

@@

\begin{itemize}
  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
  		\Comment{number of enqueue, dequeue operations in \fld{blocks} array up to this block (inclusive)}
  	\item \typ{int} \fld{super}
  		\Comment{approximate index of the block's superblock in \fld{parent.blocks}}
	\item[\com] Blocks in internal \nodes\ have the following additional fields
	\item \typ{int} \eleft, \eright
  		\Comment{index of last direct subblock in the left and right child}
  	\item[\com] Blocks in leaf \nodes\ have the following additional field
  	\item \typ{Object} \fld{element}
  		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
	\item[\com] Blocks in the root \node\ have the following additional field
	\item \typ{int} \size%
  		\Comment{size of the queue after performing all operations up to the end of this block}
\end{itemize}

>>>>

As mentioned above, we are free to order operations within a block however we like.
We order the enqueues and dequeues separately, and put the 
operations propagated from the left child before the operations from the right child.
More formally, we inductively define sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
represented by a block $B$.
If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is that operation
and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from 
$v$'s left child
and $B^R_1,\ldots,B^R_r$ from $v$'s right child, then $B$'s operation sequences are defined by the concatenations 
\begin{equation}
E(B) = E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r) \mbox{ and }
D(B) = D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
\end{equation}
We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.

@@

%The information stored in a \block\ depends on whether it is in an internal node or a leaf.
A block in a leaf represents a single enqueue or dequeue.  The block's \fld{element} field stores the value
enqueued for an enqueue operation, or \nl\ if the operation is a dequeue.

@@

\begin{itemize}
  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
  		\Comment{number of enqueue, dequeue operations in \fld{blocks} array up to this block (inclusive)}
  	\item \typ{int} \fld{super}
  		\Comment{approximate index of the block's superblock in \fld{parent.blocks}}
	\item[\com] Blocks in internal \nodes\ have the following additional fields
	\begin{itemize}[leftmargin=3mm]
		\item[$\bullet$] \typ{int} \eleft, \eright
  		\Comment{index of last direct subblock in the left and right child}
	\end{itemize}
  	\item[\com] Blocks in leaf \nodes\ have the following additional field
  	\begin{itemize}[leftmargin=3mm]
		\item[$\bullet$] \typ{Object} \fld{element}
  		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
	\end{itemize}
	\item[\com] Blocks in the root \node\ have the following additional field
	\begin{itemize}[leftmargin=3mm]
		\item[$\bullet$] \typ{int} \size%
  		\Comment{size of the queue after performing all operations up to the end of this block}
	\end{itemize}
\end{itemize}

### 2023-01-10 podc_paper/correctness-GC.tex
minor edits

\begin{proposition}
The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
\end{proposition}
\begin{proof}
\here{this proof may need some modification since code was changed to change how split point is computed}
It follows from \Cref{RBTbound} and the fact that $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can all be done in $O(\log(p+q_{max}))$ steps.

@@

We remark that the implementation remains wait-free:  the depth of recursion in each routine is  bounded
by the height of the tree and the only loop is the counted loop in the \op{Help} routine.

>>>>

\begin{mytheorem}
The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
\end{mytheorem}
\begin{proof}
\here{this proof may need some modification since code was changed to change how split point is computed}
It follows from \Cref{RBTbound} and the fact that $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can all be done in $O(\log(p+q_{max}))$ steps.

@@

We remark that the implementation remains wait-free:  the depth of recursion in each routine is  bounded
by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
Since each operation still only does two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.

\begin{customprop}{\bf{\ref{casbound}}$'$}
Each operation performs $O(\log p)$ \op{CAS} instructions in the worst case.
\end{customprop}

### 2023-01-10 podc_paper/introduction.tex
minor edits

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21}\here{try to find an older, more canonical reference}),
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first with step complexity that is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
For ease of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
bounded space while still having amortized step complexity $O(\log^2p + \log q)$ per operation.
Both implementations use single-word CAS and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
Each process propagates its operations along the path from its leaf up to the root.
Operations in the root are ordered, 
and this order is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
Helping  ensures wait-freedom and avoids the CAS retry problem:  a process propagating its operation 
to a node $v$ helps propagate all operations from both of $v$'s children.
Explicitly storing all operations in the nodes would be too costly.
Instead, we use a novel implicit representation of sets
of operations that allows us to quickly merge two sets from the children of a node,
and quickly access any  operation in a~set.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

>>>>

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21}\here{try to find an older, more canonical reference}),
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation in a system of $p$ processes.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
whose step complexity is sublinear in $p$.
Our queue is the first with step complexity that is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
For ease of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation,
and then modify it to bound the space
while still having amortized step complexity $O(\log^2p + \log q)$ per operation.
Both versions use single-word CAS on reasonably-sized words.
Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
For the space-bounded version, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
Each process adds its operations to its leaf.
As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
and avoids the CAS retry problem.
Operations in the root are ordered, 
and this order is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
Explicitly storing all operations in the nodes would be too costly.
Instead, we use a novel implicit representation of sets
of operations that allows us to quickly merge two sets from the children of a node,
and quickly access any  operation in a~set.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2023-01-10 podc_paper/podc-submission.tex
minor edits

@@

\begin{abstract}
We introduce a novel linearizable wait-free queue implementation using single-word 
CAS instructions.
Previous lock-free queues all have amortized step complexity of $\Omega(p)$ per operation in worst-case executions, 
where $p$ is the number of processes that can access the queue.
We present a new wait-free queue that takes $O(\log p)$ steps per enqueue and $O(\log^2 p +\log q)$ steps per dequeue, where $q$ is the size of the queue.
A bounded-space version of the implementation has amortized step complexity of $O(\log^2 p +\log q)$ per operation.
\end{abstract}

>>>>

\newtheorem{innercustomprop}{Proposition}
\newenvironment{customprop}[1]
  {\renewcommand\theinnercustomprop{#1}\innercustomprop}
  {\endinnercustomprop}

@@

\begin{abstract}
We present a novel linearizable wait-free queue implementation using single-word 
CAS instructions.
Previous lock-free queue implementations from \op{CAS} all have amortized step complexity of $\Omega(p)$ per operation in worst-case executions, 
where $p$ is the number of processes that access the queue.
Our new wait-free queue  takes $O(\log p)$ steps per enqueue and $O(\log^2 p +\log q)$ steps per dequeue, where $q$ is the size of the queue.
A bounded-space version of the implementation has amortized step complexity of $O(\log^2 p +\log q)$ per operation.
\end{abstract}

### 2023-01-10 podc_paper/related.tex
minor edits

\emph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free poly-logarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
We adapt their idea of processes using
a tournament tree to agree on a linearization of
operations.

@@

\emph{Fetch\&Increment Objects.}
There are sublinear-time implementations of another object
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increments
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since a queue's state cannot be represented as succinctly
as the single-word state of a fetch\&increment object.

>>>>

\emph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free poly-logarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.

@@

\emph{Fetch\&Increment Objects.}
There are implementations of fetch\&increment objects that use a polylogarithmic number of steps per operation. \cite{ERW12,10.1007/978-3-642-41527-2_20}
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which also uses a tree structure to keep track of the fetch\&increments
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since a queue's state cannot be represented as succinctly
as the single-word state of a fetch\&increment object.
